NumPy---Array Computation for Python

### 2019-12-19 paper.tex
Add linenumbers and wordcounts
Update outline and minor rearrangement

% Packages
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}
% \usepackage[multiple]{footmisc}
\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{listings, textcomp}
\lstset{language=Python}

@@

@@

@@

\subsubsection*{Project goals / scope of library}

@@

\subsubsection*{NEPS revived and roadmap}

@@

\section*{Architecture and implementation choices}

\subsection*{Library organization}
Broadly, the NumPy library consists of the following parts:
the NumPy array data structure; the so-called \emph{universal functions};
a set of library functions for manipulating arrays and doing scientific
computation; infrastructure libraries for unit tests and Python package
building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
The \code{ndarray} and the universal functions are generally considered
the core of the library.
In the following, we give a brief summary of these components of the
library.

\paragraph{\emph{Core.}}  The \code{ndarray} data structure and the
universal functions make up the core of NumPy.

@@

\paragraph{\emph{Computing libraries.}}
NumPy provides a large library of functions for array manipulation
and scientific computing, including functions for: creating, reshaping,
concatenating, and padding arrays; searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.

@@

\paragraph{\emph{Infrastructure libraries.}} NumPy provides utilities
for writing unit tests and for building Python packages.

@@

\paragraph{\emph{F2PY.}}  The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

\subsection*{Common infrastructure}

\subsubsection*{Wheels build system}

@@

\subsubsection*{NEP 29}

\subsubsection*{NumPyDoc standard}

% recent improvements

\subsection*{Language choices}

@@

\subsection*{The \_\_array\_function\_\_ Protocol}

The central position which NumPy takes within the python ecosystem with
respect to scientific computing and numerical works,
means that a vast number of projects are build on it.
These projects are consumers of the NumPy API.

On the other side, projects are providers of a
\emph{NumPy-like API} and array objects targeting audiences with specialized needs beyond
NumPy's capabilities. Some of these projects provide:

\begin{description}
    \tightlist
    \item[SciPy's \code{scipy.sparse} and \code{pydata.sparse}:] Sparse arrays
    \item[CuPy:] Computation on the GPU
    \item[Dask:] Parallel, distributed computing
    \item[PyTorch and TensorFlow:] Deep learning frameworks.
\end{description}

@@

While, as of writing this, the interface is still considered experimental
and may be modified, it has been implemented by multiple projects already
and proven useful.
% TODO: Should likely have citation/example if I claim it was useful (based
%       mainly on Matt's statement).

@@

\subsection*{GSoD}

\subsection*{Website}

@@

\subsubsection*{Impact Now}

NumPy, while having been wildly successful as a library and array
exchange project, has now also become a standard language for expressing array
and tensor computations.  For example, the NumPy API is implemented by
several popular tensor computation libraries including
CuPy\footnote{\url{https://cupy.chainer.org/}},
Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
and to a lesser extent
TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.

Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

In order to aid better interaction between NumPy and these libraries,
an ``array function'' protocol was added to NumPy in 2018 (see Section
X). The new protocol allows for external array objects (sometimes
called ``duck arrays'') to pass through NumPy:

\begin{lstlisting}
  import numpy as np
  import cupy as cp

  x_gpu = cp.array([1, 2, 3])
  y = np.sum(x_gpu)  # This works, and returns a GPU array
\end{lstlisting}

\subsubsection*{Future Development}

@@

>>>>

% Packages
\usepackage{pifont}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{float}
% \usepackage[multiple]{footmisc}
\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{moreverb} % for verbatim ouput (for wordcount)
\usepackage{listings, textcomp}
\lstset{language=Python}

@@

\usepackage{lineno}
\linenumbers

@@

% Count of words
\verbatiminput{wordcount.tex}

@@

% \subsubsection*{Project goals / scope of library}

@@

% \subsubsection*{NEPS revived and roadmap}

@@

NumPy, while having been wildly successful as a library and array exchange
project, has now also become a standard language for expressing array and
tensor computations.

\section*{Library organization}

Broadly, the NumPy library consists of the following parts:
the NumPy array data structure; the so-called \emph{universal functions};
a set of library functions for manipulating arrays and doing scientific
computation; infrastructure libraries for unit tests and Python package
building; and the program \code{f2py} for wrapping Fortran code in Python \cite{peterson2009f2py}.
The \code{ndarray} and the universal functions are generally considered
the core of the library.
In the following, we give a brief summary of these components of the
library.

\subsection*{Core.}

The \code{ndarray} data structure and the universal functions make up the core
of NumPy.

@@

\subsection*{Computing libraries}

NumPy provides a large library of functions for array manipulation
and scientific computing, including functions for: creating, reshaping,
concatenating, and padding arrays; searching, sorting and counting data
in arrays; computing elementary statistics, such as the mean, median,
variance, and standard deviation; file I/O; and more.

@@

\subsection*{Infrastructure libraries}

NumPy provides utilities for writing unit tests and for building Python
packages.

@@

The program \code{f2py} is a tool for
building NumPy-aware Python wrappers of Fortran functions.
NumPy itself does not use any Fortran code;  F2PY is part of NumPy
for historical reasons.

\section*{Common infrastructure}

In addition to the testing and building packages described above,
as the fundamental package for numerical and scientific computing
in Python, NumPy also plays a central role in building and
standardizing much of the community infrastructure.
The docstring standard mentioned in the history section is now
widely adopted.

\subsection*{Wheels build system}

@@

@@

\subsection*{The Array Function Protocol}

The central position which NumPy takes within the numerical and scientific
Python ecosystem  means that a vast number of projects are built on it.
These projects are consumers of the NumPy API.

On the other side, projects are providers of a \emph{NumPy-like API} and array
objects targeting audiences with specialized needs beyond NumPy's capabilities.
For example, the NumPy API is implemented by
several popular tensor computation libraries including
CuPy\footnote{\url{https://cupy.chainer.org/}},
Jax\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}, and
Apache MXNet\footnote{\url{https://mxnet.incubator.apache.org/api/python/docs/api/ndarray/index.html}},
and to a lesser extent
TensorFlow\footnote{\url{https://www.tensorflow.org/api\_docs/python/tf/Tensor}}.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

@@

%In order to aid better interaction between NumPy and these libraries,
%an ``array function'' protocol was added to NumPy in 2018 (see Section
%X). The new protocol allows for external array objects (sometimes
%called ``duck arrays'') to pass through NumPy:
%
%\begin{lstlisting}
%  import numpy as np
%  import cupy as cp
%
%  x_gpu = cp.array([1, 2, 3])
%  y = np.sum(x_gpu)  # This works, and returns a GPU array
%\end{lstlisting}

@@

% \subsection*{GSoD}

@@

Most of the technical work for the Moore and Sloan grant is now complete.
The main remaining item is an overhaul of the data type system, which is discussed
below.  Now that years of technical debt have been addressed, the project
is in good shape for growing the community and better addressing the needs
of new users.  That is the main focus of the CZI grant, which we discuss
below.

% \subsubsection*{Future Development}

@@

% outline plan for the CZI grant and an invitation to new contributors

%\subsection*{Website}
